Staged Abstract Interpreters

### 2019-04-06 paper/10_intro.tex
tweak figure
revise intro

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close
connection between compilers and interpreters. The first Futamura projection
specifically shows that specializing an interpreter with respect to the input
program yields an equivalent executable. Partial evaluation
\cite{DBLP:books/daglib/0072559} was the first proposed approach to realize
Futamura projections: it identifies the binding-time of variables in
the program (i.e., they can be known statically or dynamically),
evaluates the static part, and finally generates a residual program that solely
relies on the dynamic part. However, given an arbitrary program, precisely
analyzing its binding-times is hard in general. As an alternative and pragmatic
approach to specialization and partial evaluation, multi-stage programming (MSP) 
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the staging
annotations (i.e., binding-time annotations) to be given explicit from the
programmers. The staging annotations can be either syntactic (as in
MetaML/MetaOCaml, through quasi-quote, escape, etc.) or type-based \cite{DBLP:conf/gpce/RompfO10}.
These staging annotations identify which part of the inputs should be
specialized. As a classical example, we use the power function and type-based
annotations to introduce the idea of MSP:

@@

where $e$ is the program of type @Expr@ and @arg@ is the input to program $e$,
and @eval@ produces a value.
Semantically, the interpreter satisfies $ \texttt{eval}(e)(arg) = [\![ e ]\!] arg$ for all
programs and inputs. Given a program $e_0$, by applying the first Futamura
projection, we may obtain a specialized interpreter
$\texttt{eval}_{\texttt{e0}}$ with respect to $e_0$ -- the so-called \textit{equivalent executable}. 
By definition of the interpreter, 
$\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $. If we generalize the
argument @arg@ to an environment (as well as an accompanying heap object to
support mutable variables) that maps free variables in $e$ to some
values, the interpreter is a standard environment-passing interpreter,
where the specialization through MSP is also applicable.

@@

Now that we have seen two orthogonal transformations of interpreters: by
specialization, we can refactor interpreters to code generators; by
approximation, we can refactor interpreters to static analyzers. As an
intellectual quest, we should naturally ask -- starting from an abstract
interpreter, \textit{can we derive a sound abstract interpreter while it
produces optimized code that does the analysis?}
Conceptually, it is similar to the concrete setting: for an
abstract interpreter $\widehat{\texttt{eval}}: \texttt{Expr} \to
\widehat{\texttt{Env}} \to \widehat{\texttt{Value}}$, which takes a program, an abstract
environment and returns abstract values, we would like to specialize it with
respect to some program $e_0$ and produce a compiled analysis
$\widehat{\texttt{eval}}_{\texttt{e0}} : \widehat{\texttt{Env}} \to
\widehat{\texttt{Value}}$, such that $
\widehat{\texttt{eval}}_{\texttt{e0}}(\widehat{\rho}) =
\widehat{\texttt{eval}}(e_0)(\widehat{\rho})$, but the compiled one leaves the
abstract environment $\widehat{\rho}$ as remaining input.

In this paper, we study the confluence of two old ideas --- Futamura projection
and abstract interpretation, but from the perspective of their very recent
realizations --- multi-stage programming and definitional abstract interpreters.
To be specific, we present the application of the first Futamura projection on
definitional abstract interpreters, which intersects as staged abstract
interpreters. We develop a unified approach to fulfill the four different
semantics, as shown in Figure~\ref{confluence}. Starting with a generic but
parametric interpreter shared by the four semantics, by simply instantiating
the interpreter differently, we can automatically derive a concrete interpreter,
or an abstract interpreter, or their code-generator versions, respectively.

@@

\begin{itemize}
\item Env-store-value abstractions, which control whether the interpreter
  behaves as a concrete interpreter or an abstract interpreter.
  The environments, stores, and values are initially declared as abstract types in the generic interpreter.
  As shown by the AAM approach \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10}, if we widen the values to abstract domains (such as
  intervals) and refactor the store to a mapping from finite addresses to sets of
  values, we can derive a computable and sound abstract interpreter.
\item Binding-time abstractions, which control whether the interpreter produces
  values or generates code. The binding-times differentiate whether data are 
  known statically or dynamically. Unsurprisingly, such binding-time
  information can be encoded into higher-kinded types 
    \cite{DBLP:journals/jfp/CaretteKS09, Ofenbeck:2017:SGP:3136040.3136060}.
\item Monadic abstractions, which encapsulate the above two abstractions.
  The interpreter itself is written in monadic form, and by plugging different monads, 
  the underlying semantics of the interpreter is changed.
  %\todo{mention that direct-style also works?}
  %\cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95, DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979}
\end{itemize}

We show that these well-known abstractions can be combined smoothly, and more
importantly for free. We implement the four interpreters using Scala and the
Lightweight Modular Staging \cite{DBLP:conf/gpce/RompfO10} framework. Through
staging, the major overhead of abstractions -- from the monadic layers -- is
completely eliminated in the generated code. We observe that the staged version is 
11x times as fast as the unstage abstract interpreter. 
However, we do not lose any advantages of this high-level programming style 
(e.g., modularity and extensibility); the meaning of the program also remains 
the same, particularly, the soundness of static analysis after staging.

%\paragraph{Performance via Specialization}
Besides the intellectual merit shown in the confluence diagram, this paper
contributes to a practical problem in static analysis: \textit{can we speed up a
static analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and
\textit{efficiency}, we argue that applying meta-programming (multi-stage
programming, in particular) with abstract interpreters is an effective approach
to improve the latter while the former is untouched.
In the literature, specializing analysis with respect to the input program
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587,
ashley:practical} is not new, despite the fact that most of the previous works
are ad-hoc on a particular analysis or requiring significant changes to the
analyzer. For example, abstract compilation is one of the techniques
\cite{Boucher:1996:ACN:647473.727587}. But it requires the whole analyzer to be
rewritten to the closure generation form. In this paper, we advocate a systematic
approach of meta-programming that achieves the performance goal with less
intrusive changes and engineering efforts.

@@

Furthermore, we show various applications of staging an abstract interpreter.
(1) We compare our approach with abstract compilation, and show that we can
achieve the same optimization by utilizing type-based stage annotations,
meanwhile, the analyzer program does not need any changes. (2) We extend the basic staged abstract interpreter to diverse control-flow analysis, including store-widening analysis, context-sensitive analysis and abstract garbage collection.
(3) Staging abstract interpreter enables modular analysis and provides a mechanized
way to create library summaries. We notice that we may apply the staged abstract
interpreter separately on the different part of the analyzed program, the
generated code is reusable and composable. Therefore, we can turn a
whole-program analyzer into an analyzer that runs modularly.

@@

\paragraph{Contributions} Briefly, the contribution of the paper is twofold:
\begin{enumerate}[leftmargin=2em]
  \item Intellectually, through multi-stage programming, we present a unified construction 
    that naturally extends the first Futamura projection to abstract interpreters.
 \item Pragmatically, by case studies and empirical evaluation, we show that staging 
   abstract interpreters is useful to improve performance while untouching
   the soundness. 
\end{enumerate}
\paragraph{Organization} The paper is organized as follows:
\begin{itemize}[leftmargin=2em]
\item Starting from a generic interface of the interpreter for a small higher-order language
  (Section~\ref{prelim}), we first show the instantiation of concrete
  interpretation (Section~\ref{unstaged_conc}), and then lift it to the staged
  version by replacing the binding-time types (Section~\ref{stagedinterp}).
\item We discuss the unstaged abstract interpreter under the same framework by
  replacing the environment, store and values (Section~\ref{unstaged_abs}).
  Finally, we show the combination of approximation and specialization, dubbed
  \textit{staged abstract interpreters}, can be readily derived (Section~\ref{sai}).
\item We conduct three case studies on the applicability of our approach
  (Section~\ref{cases_study}).
  \begin{enumerate}
  \item We first revisit the Abstract Compilation (AC)
    \cite{Boucher:1996:ACN:647473.727587} (Section~\ref{cs_ac}), showing that we can
    achieve the same optimization with much less engineering effort.
  \item We extend the staged abstract interpreter to various control-flow
    analysis (Section \ref{cfa}), including context-sensitivity, store-widening, and abstract
    garbage collection, showing that our approach is widely applicable.
  \item We apply our approach to modular analysis (Section \ref{modular}),
    i.e., analyzing the components of the program separately and composing them later,
    showing a novel application of generative programming with program analysis.
  \end{enumerate}
\item We empirically evaluate the performance improvement by staging on
    control-flow analysis (Section~\ref{evaluation}). We compare the both
    context-insensitive and store-widening analysis.
\end{itemize}

>>>>

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close
connection between compilers and interpreters. The first Futamura projection
specifically shows that specializing an interpreter with respect to the input
program yields an equivalent executable. Partial evaluation
\cite{DBLP:books/daglib/0072559} was the first proposed approach to realize
Futamura projections: it identifies the binding-time of variables in
the program (i.e., they can be known statically or dynamically),
evaluates the static part, and finally generates a residual program that solely
relies on the dynamic part. However, given an arbitrary program, precisely
analyzing its binding-times is hard in general. As an alternative and pragmatic
approach to specialization and partial evaluation, multi-stage programming (MSP) 
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the staging
annotations (i.e., binding-time annotations) to be given explicitly from the
programmers. The staging annotations can be either syntactic (as in
MetaML/MetaOCaml, through quasi-quote, escape, etc.) or type-based \cite{DBLP:conf/gpce/RompfO10}.
These staging annotations identify which part of the inputs should be
specialized. As a classical example, we use the power function and type-based
annotations to introduce the idea of MSP:

@@

where $e$ is a program of type @Expr@ and @arg@ is the input to $e$,
and @eval@ produces a value.
Semantically, the interpreter satisfies $ \texttt{eval}(e)(arg) = [\![ e ]\!] arg$ for all
programs and inputs. Given a program $e_0$, by applying the first Futamura
projection, we can obtain a specialized interpreter
$\texttt{eval}_{\texttt{e0}}$ with respect to $e_0$, which is the so-called \textit{equivalent executable}. 
By definition of the interpreter, 
$\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $. If we generalize the
argument @arg@ to an environment that maps free variables in $e$ to some
values, the interpreter is a standard environment-passing interpreter,
where the specialization through MSP is also applicable.

@@

Now that we have seen two orthogonal transformations of interpreters: by
specialization, we can refactor interpreters to code generators; by
approximation, we can refactor interpreters to static analyzers. As an
intellectual quest, we should naturally ask -- starting from an abstract
interpreter, \textit{can we derive a sound abstract interpreter while it
produces optimized code that does the analysis?}
Conceptually, it is similar to the concrete setting: given an
abstract interpreter $\widehat{\texttt{eval}}: \texttt{Expr} \to
\widehat{\texttt{Env}} \to \widehat{\texttt{Value}}$, which takes a program, an abstract
environment and returns abstract values, we would like to specialize it with
respect to some program $e_0$ and produce a compiled analysis
$\widehat{\texttt{eval}}_{\texttt{e0}} : \widehat{\texttt{Env}} \to
\widehat{\texttt{Value}}$, where the abstract environment $\widehat{\rho}$ is kept as input.
The specialized analysis satisfies $\widehat{\texttt{eval}}_{\texttt{e0}}(\widehat{\rho}) =
\widehat{\texttt{eval}}(e_0)(\widehat{\rho})$.

In this paper, we study the confluence of two old ideas --- Futamura projection
and abstract interpretation, but from the perspective of their very recent
realizations --- multi-stage programming and definitional abstract interpreters.
To be specific, we present the application of the first Futamura projection on
definitional abstract interpreters, which intersects as staged abstract
interpreters. We develop a unified approach to fulfill the four different
semantics, as shown in Figure~\ref{confluence}. Starting with a generic but
parametric interpreter shared by the four semantics, by simply instantiating
the interpreter differently, we can automatically derive a concrete interpreter,
or an abstract interpreter, or their staged versions that generate code, respectively.

@@

\begin{itemize}
\item Env-store-value abstractions, which define the domains that the interpreter will be running with.
  The environments, stores, and values are initially declared as abstract types in the generic interpreter.
  As shown by the AAM approach \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10}, if we widen the values to abstract domains (such as
  intervals) and refactor the store to a mapping from finite addresses to sets of
  values, we can derive a computable and sound abstract interpreter.
\item Binding-time abstractions, which control whether the interpreter produces
  values or generates code. The binding-times differentiate whether data are 
  known statically or dynamically. Unsurprisingly, such binding-time
  information can be encoded into higher-kinded types 
    \cite{DBLP:journals/jfp/CaretteKS09, Ofenbeck:2017:SGP:3136040.3136060}.
\item Monadic abstractions, which encapsulate the above two abstractions.
  The interpreter itself is written in monadic form, and by plugging different monads, 
  the underlying semantics of the interpreter is changed. We also use monad transformers to combine multiple monadic effects.
  %\todo{mention that direct-style also works?}
  %\cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95, DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979}
\end{itemize}

We show that these well-known abstractions can be combined smoothly, and more
importantly for free. We implement the four interpreters using Scala and the
Lightweight Modular Staging \cite{DBLP:conf/gpce/RompfO10} framework. Through
staging, the major overhead of abstractions -- from the monadic layers -- is
completely eliminated in the generated code. We observe that the staged version is 
on average 11x times as fast as the unstaged abstract interpreter. 
However, we do not lose any advantages of this high-level programming style,
e.g., modularity and extensibility. The meaning of generated programs also remain
the same, particularly, the soundness of static analysis after staging.

%\paragraph{Performance via Specialization}
Besides the intellectual merit shown in the confluence diagram, this paper
contributes to a practical problem in static analysis: \textit{can we speed up a
static analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and
\textit{efficiency}, we argue that applying meta-programming (multi-stage
programming, in particular) with abstract interpreters is an effective approach
to improve the latter while the former is untouched.
In the literature, specializing analysis with respect to the input program
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587,
ashley:practical} is not new, despite the fact that most of the previous works
are ad-hoc on a particular analysis or requiring significant changes to the
analyzer. For example, abstract compilation is one of the techniques
\cite{Boucher:1996:ACN:647473.727587}. But it requires the whole analyzer to be
rewritten to the closure generation form. In this paper, we advocate a systematic
approach of generative programming that achieves the performance goal with less
intrusive changes and engineering efforts.

@@

Furthermore, we show various applications of staging an abstract interpreter.
(1) We compare our approach with abstract compilation, and show that we can
achieve the same optimization by utilizing type-based stage annotations,
meanwhile, the analyzer program does not need any changes. (2) We extend the 
basic staged abstract interpreter to diverse control-flow analysis, 
including a store-widened analysis, a context-sensitive analysis and abstract garbage collection.
(3) Staging abstract interpreter enables modular analysis and provides a mechanized
way to create library summaries. We notice that we may apply the staged abstract
interpreter separately on the different part of the analyzed program, and the
generated code is reusable and composable. Therefore, we can turn a
whole-program analyzer into an analyzer that runs modularly.

@@

\paragraph{Contributions} Briefly, the contribution of the paper is twofold:
\begin{enumerate}[leftmargin=2em]
  \item Intellectually, through multi-stage programming, we present a unified construction 
    that naturally extends the first Futamura projection to abstract interpreters.
 \item Pragmatically, by case studies and empirical evaluation, we show that staging 
   abstract interpreters is useful to improve performance while untouching
   the soundness. 
\end{enumerate}
\paragraph{Organization} The paper is organized as follows:
\begin{itemize}[leftmargin=2em]
\item Starting from a generic interface of the interpreter for a small higher-order language
  (Section~\ref{prelim}), we first show the instantiation of concrete
  interpretation (Section~\ref{unstaged_conc}), and then lift it to the staged
  version by replacing the binding-time types (Section~\ref{stagedinterp}).
\item We present the unstaged abstract interpreter under the same framework by
  replacing the environment, store and values (Section~\ref{unstaged_abs}).
  Finally, we show the combination of approximation and specialization, dubbed
  \textit{staged abstract interpreters}, can be readily derived (Section~\ref{sai}).
\item We conduct three case studies on the applicability of our approach
  (Section~\ref{cases_study}).
  \begin{enumerate}
  \item We first revisit the Abstract Compilation (AC)
    \cite{Boucher:1996:ACN:647473.727587} (Section~\ref{cs_ac}), showing that we can
    achieve the same optimization with much less engineering effort.
  \item We extend the staged abstract interpreter to various control-flow
    analysis (Section \ref{cfa}), including context-sensitivity, store-widening, and abstract
    garbage collection, showing that our approach is widely applicable.
  \item We apply our approach to modular analysis (Section \ref{modular}),
    i.e., analyzing the components of the program separately and composing them later,
    showing a novel application of generative programming with program analysis.
  \end{enumerate}
\item We empirically evaluate the performance improvement by staging on
    control-flow analysis (Section~\ref{evaluation}). We compare the both
    context-insensitive and store-widening analysis.
\end{itemize}

### 2019-04-06 paper/abstract.tex
revise intro

As an example of abstraction without regret, we show that staging abstract
interpreters is practical and useful to optimize static analysis while requiring
least engineering efforts and not compromising any soundness. We conduct three
case studies, including a comparison with
\citeauthor{Boucher:1996:ACN:647473.727587}'s abstract compilation,
the application of various control-flow analyses, and the use to modular
analysis. We also empirically evaluate the performance improved by staging. The
overhead of the abstraction layers is eliminated in the generated code, and the 
experiment shows an average speedup of 7x with staging for control-flow analysis.

>>>>

As an example of abstraction without regret, we show that staging abstract
interpreters is practical and useful to optimize static analysis while requiring
least engineering efforts and not compromising any soundness. We conduct three
case studies, including a comparison with
\citeauthor{Boucher:1996:ACN:647473.727587}'s abstract compilation,
the application on various control-flow analyses, and the use to modular
analysis. We also empirically evaluate the performance improved by staging. The
overhead of the abstraction layers is eliminated in the generated code, and the 
experiment shows an average speedup of 11x times with staging for control-flow analysis.
